---
title: "LIMMA"
author: "gideonsamuel"
date: "2025-09-13"
output: html_document
---
#Load Required Libraries
```{r}
library(GEOquery)
library(limma)
library(matrixStats)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(ggrepel)
library(enrichplot)
library(pheatmap)
```

```{r}
dir.create("results", showWarnings = FALSE)
dir.create("results/figures", showWarnings = FALSE)
dir.create("results/tables", showWarnings = FALSE)
```


#Data Acquisition
```{r}
id <- "GSE19188"
e_set <- getGEO(id, GSEMatrix = TRUE)
e_set <- e_set[[1]]

expn <- exprs(e_set)
pheno <- pData(e_set)
```

#Initial Data Inspection
```{r}
dim(expn)
head(expn[,1:5])
any(is.na(expn))
head(pheno[,1:20])
```

#Data Normalization
```{r}
range(expn)#range before normalization
# normalize (GEO data often normalized, but included for consistency)
expn <- normalizeBetweenArrays(expn)
range(expn)#range after normalization
```

#Sample Group Definition
```{r}
group <- ifelse(grepl("tissue type: tumor", pheno$characteristics_ch1), "Disease",
         ifelse(grepl("tissue type: healthy", pheno$characteristics_ch1), "Control", NA))

group <- factor(group, levels = c("Control", "Disease"))
table(group)
```

#Data filtering and preprocessing
```{r}
# keep only samples with a label
keep <- !is.na(group)

expn_sub <- expn[, keep]
group <- group[keep]

all(colnames(expn_sub) == rownames(pheno)[keep])  # should be TRUE
keep_genes <- rowVars(expn_sub) > 0
expn_sub <- expn_sub[keep_genes, ]
sum(keep_genes)
```

#Quality Control: PCA
```{r}
pca <- prcomp(t(expn_sub), scale. = TRUE)

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  Group = group
)

pca_summary <- summary(pca)
var <- pca_summary$importance[2,]

p <- ggplot(pca_df, aes(PC1, PC2, color = Group)) +
  geom_point(size = 3) +
  stat_ellipse() +
  labs(
    x = paste0("PC1 (", round(var[1]*100, 2), "%)"),
    y = paste0("PC2 (", round(var[2]*100, 2), "%)")
  ) +
  theme_minimal() +
  theme(text = element_text(size = 14))

p

ggsave("results/figures/pca.png", p, width = 6, height = 5, dpi = 300)
```

#Differential expression analysis (LIMMA)
```{r}
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)

fit <- lmFit(expn_sub, design)

contrast.matrix <- makeContrasts(Disease-Control, levels = design)

fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

result <- topTable(fit2, number = Inf, adjust = "fdr")
```

#Summary of Differential Expression Results
```{r}
cat("Significant genes (FDR < 0.05 & |logFC| > 1):",
    sum(result$adj.P.Val < 0.05 & abs(result$logFC) > 1), "\n")
head(result[, c("logFC", "adj.P.Val")], 10)
```

#Significant Gene Extraction
```{r}
# Significant genes (FDR < 0.05 and |logFC| > 1)

sig_genes <- result[result$adj.P.Val < 0.05 & abs(result$logFC) > 1, ]

cat("Number of significant genes:", nrow(sig_genes), "\n")

# Top upregulated genes
head(sig_genes[order(-sig_genes$logFC), ], 10)

# Top downregulated genes
head(sig_genes[order(sig_genes$logFC), ], 10)
```

#Gene Annotation and Mapping
```{r}
gpl <- getGEO(annotation(e_set))
annot <- Table(gpl)

# Add probe IDs
result$ProbeID <- rownames(result)

# Merge with annotation
res_annot <- merge(result, annot, by.x = "ProbeID", by.y = "ID")

# Clean gene symbols
res_annot <- res_annot[!is.na(res_annot$`Gene Symbol`) &
                       res_annot$`Gene Symbol` != "", ]

# Sort by significance FIRST
res_annot <- res_annot[order(res_annot$adj.P.Val), ]

# Remove duplicates AFTER sorting (keep best probe per gene)
res_annot <- res_annot[!duplicated(res_annot$`Gene Symbol`), ]
```

#Expression Matrix Conversion (Probe to Gene)
```{r}
#Convert expression matrix to gene symbols
mapping <- res_annot[, c("ProbeID", "Gene Symbol")]

# Keep only probes present in expression matrix
mapping <- mapping[mapping$ProbeID %in% rownames(expn_sub), ]

# Match order with expression matrix
mapping <- mapping[match(rownames(expn_sub), mapping$ProbeID), ]

# Remove NA matches
valid <- !is.na(mapping$ProbeID)

expn_gene <- expn_sub[valid, ]
mapping   <- mapping[valid, ]

# Assign gene symbols as rownames
rownames(expn_gene) <- mapping$`Gene Symbol`

# Remove any accidental duplicate gene names
expn_gene <- expn_gene[!duplicated(rownames(expn_gene)), ]
```

#Annotated Significant Genes
```{r}
#Significant genes

sig_annot <- res_annot[res_annot$adj.P.Val < 0.05 & abs(res_annot$logFC) > 1, ]

#Volcano Plot gene selection
top_up   <- sig_annot[sig_annot$logFC > 1, ]
top_down <- sig_annot[sig_annot$logFC < -1, ]

top_up   <- head(top_up[order(top_up$adj.P.Val), ], 5)
top_down <- head(top_down[order(top_down$adj.P.Val), ], 5)

top_genes_volcano <- rbind(top_up, top_down)
```

#Volcano plot Visualization
```{r}
# Define significance
result$Significance <- "Not Significant"
result$Significance[result$adj.P.Val < 0.05 & result$logFC > 1] <- "Up"
result$Significance[result$adj.P.Val < 0.05 & result$logFC < -1] <- "Down"

# Count genes
n_up   <- sum(result$Significance == "Up")
n_down <- sum(result$Significance == "Down")

# Plot
v <- ggplot(result, aes(x = logFC, y = -log10(adj.P.Val), color = Significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  
  scale_color_manual(values = c(
    "Down" = "blue",
    "Not Significant" = "grey",
    "Up" = "red"
  )) +
  
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +

  # Smart labels
  geom_text_repel(
  data = top_genes_volcano,
  aes(x = logFC, 
      y = -log10(adj.P.Val), 
      label = `Gene Symbol`,
      color = Significance
      ),
    inherit.aes = FALSE,
    size = 3
    )+
  
  labs(
    title = paste0("Volcano Plot (Up: ", n_up, ", Down: ", n_down, ")"),
    x = "log2 Fold Change",
    y = "-log10 Adjusted P-value"
    ) +
  theme_minimal() +
  theme(text = element_text(size = 14))

v

ggsave("results/figures/volcano.png", v, width = 6, height = 5, dpi = 300)
```

#Functional Enrichment Analysis
```{r}
genes <- unique(sig_annot$`Gene Symbol`)
genes <- genes[!is.na(genes)]

gene_df <- bitr(genes,
                fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Hs.eg.db)

entrez_ids <- unique(gene_df$ENTREZID)

if (length(entrez_ids) == 0) {
  stop("No genes mapped to ENTREZ IDs")
}
```

#GO Enrichment Analysis
```{r}
ego <- enrichGO(gene          = entrez_ids,
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05)

head(ego)
# GO enrichment visualization
g <- dotplot(ego, showCategory = 10) +
  scale_color_gradient(low = "#2c7bb6", high = "#d7191c") +
  labs(title = "GO Biological Processes") +
  theme_minimal() +
  theme(text = element_text(size = 14))

g

ggsave("results/figures/go.png", g, width = 7, height = 5, dpi = 300)
```

#KEGG Pathway Enrichment
```{r}
kegg <- enrichKEGG(gene         = entrez_ids,
                   organism     = "hsa",
                   pvalueCutoff = 0.05)

head(kegg)

# KEGG enrichment visualization
k <- dotplot(kegg, showCategory = 10) +
  scale_color_gradient(low = "#2c7bb6", high = "#d7191c") +
  labs(title = "KEGG Pathways") +
  theme_minimal() +
  theme(text = element_text(size = 14))

k

ggsave("results/figures/kegg.png", k, width = 7, height = 5, dpi = 300)
```

#Heatmap of Top Differentially Expressed Genes
```{r, fig.width=10, fig.height=8}
annotation_col <- data.frame(Group = group)
rownames(annotation_col) <- colnames(expn_sub)

# Select top genes by BOTH significance and variability
top_genes_heatmap <- head(sig_annot[order(abs(sig_annot$logFC), decreasing = TRUE), ], 50)

gene_names <- top_genes_heatmap$`Gene Symbol`

valid_genes <- gene_names[gene_names %in% rownames(expn_gene)]

if (length(valid_genes) == 0) {
  stop("No valid genes found for heatmap")
}

mat <- expn_gene[valid_genes, , drop = FALSE]

pheatmap(mat,
         scale = "row",
         annotation_col = annotation_col,
         show_colnames = FALSE,
         show_rownames = TRUE,
         clustering_method = "complete",
         main = "Top Differentially Expressed Genes",
         fontsize = 6,
         filename = "results/figures/heatmap_genes.png",
         width = 8,
         height = 10)
```

#Sample Clustering (Distance Heatmap)
```{r, fig.width=10, fig.height=8}
dist_mat <- dist(t(expn_sub))
hc <- hclust(dist_mat)

pheatmap(as.matrix(dist_mat)[hc$order, hc$order],
         annotation_col = annotation_col,
         show_colnames = FALSE,
         show_rownames = FALSE,
         clustering_method = "complete",
         main = "Sample Distance Heatmap",
         fontsize = 6,
         filename = "results/figures/heatmap_samples.png",
         width = 8,
         height = 10)
```

#Final Gene Inspection
```{r}
# Top significant genes
head(res_annot[order(res_annot$adj.P.Val),
               c("Gene Symbol", "logFC", "adj.P.Val")], 20)

# Top upregulated genes
head(res_annot[order(-res_annot$logFC),
               c("Gene Symbol", "logFC", "adj.P.Val")], 20)
```

# Save results
```{r}
write.csv(sig_genes, "results/tables/DE_genes.csv", row.names = TRUE)
write.csv(as.data.frame(ego), "results/tables/GO_results.csv", row.names = FALSE)
write.csv(as.data.frame(kegg), "results/tables/KEGG_results.csv", row.names = FALSE)
```